---
title: "Maths and Stats Coursework"
author: "Georgina Adams"
date: "07/11/25"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Question 1: Simulating a Board Game

The original `play_mc()` function contained a bug in its end-of-game check:
it set `pos` to 70 (ending the game) whenever the position was **below** 70,
rather than when it reached or exceeded 70. The corrected version below fixes
this logic and extends the game with slippery squares (which slide the
player back) and climbing squares (which move the player forward).

```{r}
# Load libraries
source("Functions_for_CW.R")
library(dplyr)
library(ggplot2)
library(tidyr)

# Original (buggy) version of the game function:
# this immediately assigns pos as 70 if pos is currently below 70,
# ending the game far too early.
play_mc <- function(start_pos = 0) {
  die <- sample.int(6, 1)
  pos <- start_pos
  pos <- pos + die

  if (pos < 70) {
    pos <- 70
    return(pos)
  }
}

# Corrected and extended version:
play_mc <- function(start_pos = 0, slippery_squares = c(9, 18, 27, 36, 45, 54, 63), climbing_squares = c(16, 32, 48, 64)) {
  die <- sample.int(6, 1)
  pos <- start_pos
  pos <- pos + die

  # Game now correctly ends once the score reaches or exceeds 70.
  if (pos >= 70) {
    pos <- 70
    return(pos)
  }

  # If the player lands on a slippery square, they slide back 1-3 places
  # (never below 0).
  if (pos %in% slippery_squares) {
    slide <- sample(1:3, 1)
    pos <- max(0, pos - slide)
  }

  # If the player lands on a climbing square, they move forward 5 places
  # (never above 70).
  else if (pos %in% climbing_squares) {
    pos <- min(70, pos + 5)
  }

  return(pos)
}

play_mc()
```

# Question 2: Simulating Full Games

```{r}
play_mc_full()
```

# Question 3: Distribution of Turns to Finish

```{r}
results_sc <- simulate_climbs()

my_hist <- hist(results_sc$turns_vector,
     main = "Histogram showing the frequency of turns taken to finish each game",
     xlab = "Number of turns",
     ylab = "Frequency")
```

# Question 4: Position After a Fixed Number of Turns

```{r}
results_sp <- simulate_pos_at()

n_turn_hist <- hist(results_sp$positions_at_n_turns,
     main = "Histogram showing positions at 10 turns taken",
     xlab = "Position at 10 turns",
     ylab = "Frequency")

# Re-run with an updated slip-back range and climb-up amount
results_sp_new_values <- simulate_pos_at(slip_back_range = 4:8, climb_up_amount = 2)

new_values_hist <- hist(results_sp_new_values$positions_at_n_turns,
     main = "Histogram showing positions at 10 turns with updated values",
     xlab = "Position at 10 turns",
     ylab = "Frequency")

print(results_sp$turns_stats)
print(results_sp_new_values$turns_stats)
```

For the first simulation we have a mean position of around 32 at 10 turns,
whereas after changing the values and re-running the simulations with the
updated slip-back range and climb-up amount, we have a mean position of
around 27. This is expected based on the changes, as the position slips
back further and climbs up less when landing on slippery and climbing
squares. This shift is demonstrated in the histograms shown above.

# Question 5: BMI, Density Curves, and PCA

```{r}
# The data file is not located in a directory called "data", so the
# originally-supplied load() path did not work; corrected below.
load("assess_data_1225.Rdata")

dens_m <- density(bmi_data$bmi[bmi_data$sex == "male"])
dens_f <- density(bmi_data$bmi[bmi_data$sex == "female"])

plot(dens_f, col = "green", main = "BMI Density Plot of Male and Female Participants", xlab = "BMI", ylab = "Density")
lines(dens_m)
legend("topright", c("Female", "Male"), col = c("green", "black"), lty = 1)

idx <- 1:200

# Runs PCA on a subset of the data (columns 1 and 9 removed).
pca_res <- prcomp(bmi_data[idx, -c(1, 9)], center = TRUE, scale = TRUE)

colours <- ifelse(bmi_data$sex == "male", "blue", "pink")

plot(pca_res$x[, 1], pca_res$x[, 2], col = colours, pch = 19, main = "Principal Component Analysis of BMI data subset", xlab = "Principle component 1", ylab = "Principle component 2")

# The originally-supplied model did not reference the data frame correctly;
# corrected below.
out <- lm(bmi_data$bmi[idx] ~ bmi_data$age[idx])
summary(out)

p_val <- summary(out)$coefficients[2, 4]
p_val
```

# Question 6: Data Quality and BMI Correction

```{r}
# Summary of mean BMI values by sex. Both averages are far beyond the
# normal range, indicating implausible values and quality issues in the
# dataset as supplied.
aggregate(bmi ~ sex, data = bmi_data, mean)
summary(bmi_data$bmi)
```

```{r}
ggplot(bmi_data, aes(x = bmi)) +
  geom_histogram(fill = "darkblue", color = "black") +
  geom_vline(xintercept = c(20, 30), linetype = "dashed", color = "black") +
  labs(title = "BMI Distribution of Participants", x = "BMI", y = "Count")
```

The distribution is severely skewed, suggesting extreme values are pulling
it to one side and outside of normal ranges. Dashed lines mark where
normal BMI values should fall.

```{r}
# Recalculates true BMI as weight (kg) divided by height (m) squared,
# and adds it as a new column.
library(dplyr)
bmi_data <- bmi_data %>%
  mutate(Correct_bmi = weight / ((height / 100)^2))

ggplot(bmi_data, aes(x = Correct_bmi)) +
  geom_histogram(fill = "darkblue", color = "black") +
  geom_vline(xintercept = c(20, 30), linetype = "dashed", color = "black") +
  labs(title = "Distribution of Corrected BMI", x = "True BMI of Male and Female Participants", y = "Number of Participants")

ggplot(bmi_data, aes(sex, Correct_bmi, fill = sex)) +
  geom_boxplot() +
  scale_fill_manual(values = c(female = "pink", male = "blue")) +
  geom_hline(yintercept = c(20, 30), linetype = "dashed") +
  labs(title = "Boxplot of Corrected BMI", y = "True BMI", x = "Sex")
```

After correction, the histogram peak now falls within the dashed lines
showing the expected range. The boxplot shows males with a slightly higher
median BMI than females, which is physiologically normal and still within
the expected range.

```{r}
aggregate(Correct_bmi ~ sex, data = bmi_data, mean)
summary(bmi_data$Correct_bmi)
```

Mean corrected BMI is now within the normal range, but minimum/maximum
values remain abnormal and will require further quality control, including
removal of NAs.

```{r}
# PCA plot to visualise outliers using clean numerical data.
pca_input <- bmi_data[, -c(1, 9)]
pca_input <- pca_input[sapply(pca_input, is.numeric)]

pca_input <- data.frame(lapply(pca_input, function(x) replace(x, !is.finite(x), NA)))
pca_input <- na.omit(pca_input)

p <- prcomp(scale(pca_input))
pc_dataframe <- as.data.frame(p$x)
var_explained <- round(100 * p$sdev^2 / sum(p$sdev^2), 1)

# Outliers computed using Tukey's rule: a fence is set outside Q1 and Q3
# for each of the first two principal components.
Q1_PC1 <- quantile(pc_dataframe$PC1, 0.25)
Q3_PC1 <- quantile(pc_dataframe$PC1, 0.75)
IQR_PC1 <- Q3_PC1 - Q1_PC1
lower_PC1 <- Q1_PC1 - 1.5 * IQR_PC1
upper_PC1 <- Q3_PC1 + 1.5 * IQR_PC1

Q1_PC2 <- quantile(pc_dataframe$PC2, 0.25)
Q3_PC2 <- quantile(pc_dataframe$PC2, 0.75)
IQR_PC2 <- Q3_PC2 - Q1_PC2
lower_PC2 <- Q1_PC2 - 1.5 * IQR_PC2
upper_PC2 <- Q3_PC2 + 1.5 * IQR_PC2

outliers <- (pc_dataframe$PC1 < lower_PC1 | pc_dataframe$PC1 > upper_PC1) |
            (pc_dataframe$PC2 < lower_PC2 | pc_dataframe$PC2 > upper_PC2)

colours <- ifelse(bmi_data$sex == "male", "blue", "pink")

plot(pc_dataframe$PC1, pc_dataframe$PC2,
     pch = 19, cex = 0.3,
     col = colours,
     xlab = paste0("PC1 (", var_explained[1], "%)"),
     ylab = paste0("PC2 (", var_explained[2], "%)"),
     main = "PCA of BMI Data with Outliers Highlighted")

points(pc_dataframe$PC1[outliers], pc_dataframe$PC2[outliers],
       col = "red",
       pch = 19,
       cex = 0.5)

legend("topright", legend = c("Male", "Female", "Outlier"),
      col = c("blue", "pink", "red"), pch = 19)
```

The PCA plot shows two clear clusters corresponding to male and female
participants. Red points mark extreme values that remain in the dataset.
PC1 explains 55.6% of the variance across all numeric variables, and a
further 26.8% is explained by PC2 — together 84.4% of the variance, which
is enough for the trends and clusters to be shown reliably in two
dimensions.

```{r}
# Final cleaned dataset: NAs and outliers removed, implausible min/max
# values resolved. This also replaces the original bmi column.
cleaned_bmi_data <- update_bmi(bmi_data)
summary(cleaned_bmi_data)
```

# Question 7: Predicting Body Fat from BMI

Using simple linear regression to check how well BMI (covariate) can
predict body fat (response), checking model assumptions and plotting the
residuals.

```{r}
library(ggplot2)
library(reshape2)
library(bbmle)

lm_model <- lm(body_fat ~ Correct_bmi, data = cleaned_bmi_data)
summary(lm_model)

plot(lm_model)
```

The Residuals vs Fitted plot shows residuals scattered around 0, meaning
the linearity assumption is reasonably met, although funnelling toward
the right indicates heteroscedasticity (non-constant variance), and the
plot highlights some outliers. Variance also increases as fitted values
increase. The R² value of 0.2956 shows that BMI alone explains only 29.6%
of the variation in body fat — this is not a strong model on its own.

The Q-Q plot, used to check distributional assumptions, shows both tails
deviating from the theoretical distribution: a heavy tail indicating more
extreme residuals than expected under normality, and a light tail
indicating fewer extreme residuals than expected. This is most likely
influenced by outliers and by other covariates not yet accounted for.

```{r}
lr_plot <- ggplot(cleaned_bmi_data, aes(x = Correct_bmi, y = body_fat)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "blue") +
  labs(
    title = "Simple Linear Regression model: Body Fat vs BMI",
    x = "Predictor: BMI",
    y = "Response: Body Fat")

print(lr_plot)
```

This plot shows the regression line fitted to BMI vs body fat. There is a
clear linear association between the predictor and response, although BMI
alone is not a reliable predictor of body fat on its own.

```{r}
residuals_df1 <- data.frame(resid = residuals(lm_model))

ggplot(residuals_df1, aes(x = resid)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "darkblue",
                 color = "black") +
  stat_function(fun = dnorm,
                args = list(mean = mean(residuals_df1$resid, na.rm = TRUE),
                            sd = sd(residuals_df1$resid, na.rm = TRUE)),
                color = "lightblue",
                size = 1) +
  labs(title = "Histogram of Residuals with Fitted Normal Curve",
       x = "Residuals",
       y = "Density")
```

The residual histogram lacks the bell shape expected under normality, and
instead shows a bimodal distribution with two peaks and a trough around
0. This suggests two distinct subgroups in the data. Given that males tend
to have a higher BMI than females, sex is likely one of the key covariates
missing from this model, and including it should improve model fit.

## Extending the Model with Additional Covariates

This model instead predicts body fat using age, BAI, body density, weight,
height, and corrected BMI as predictors.

```{r}
lm_bodyfat_covariants <- lm(body_fat ~ age + bai + density + weight + height + Correct_bmi,
                 data = cleaned_bmi_data)
summary(lm_bodyfat_covariants)
```

The R² value of 0.9961 shows that 99.6% of the variation in body fat is
explained by these six covariates — a very strong model fit. Based on the
absolute t-values, which show how much each predictor contributes once
accounting for its variance, **density** is by far the dominant predictor
of body fat.

```{r}
plot(lm_bodyfat_covariants)
```

The Residuals vs Fitted plot shows points symmetrically distributed and
clustering toward the middle, suggesting generally constant variance with
no obvious pattern of non-linearity. Two outliers appear far from the
bulk of the data and will have some influence on the model's estimates;
however, since quality control has already been applied to the data, these
likely represent natural variability in the population rather than data
errors.

The Q-Q plot further supports a normal distribution of residuals, which
sit close to the expected quantiles, with only slight deviation at the
positive and negative tails.

```{r}
residuals_df2 <- data.frame(resid = residuals(lm_bodyfat_covariants))

ggplot(residuals_df2, aes(x = resid)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "darkblue",
                 color = "black") +
  stat_function(fun = dnorm,
                args = list(mean = mean(residuals_df2$resid, na.rm = TRUE),
                            sd = sd(residuals_df2$resid, na.rm = TRUE)),
                color = "lightblue",
                size = 1) +
  labs(title = "Histogram of Residuals with Fitted Normal Curve",
       x = "Residuals",
       y = "Density")
```

The residuals histogram shows a close-to-normal distribution, following
the fitted normal curve with clustering around the mean and median. This
supports the R² value of 0.9961, confirming this as a strong model.
