---
title: "Maths and Stats Coursework"
author: "Georgina Adams"
date: "07/11/25"
output: "html_document"
---

```{r}

#Q1

# Load libraries

source("Functions_for_CW.R")
library(dplyr)
library(ggplot2)
library(tidyr)


# First line of code defines the function for the game
play_mc <- function(start_pos = 0) {  
  
# This returns a random integer between 1 and 6
  die <- sample.int(6, 1)
  
# Assigns value start_pos to variable pos.
  pos <- start_pos
  
# Computes the movement of the player based on the integer value from the dice 
# roll and assigns it to the variable pos.
  pos <- pos + die
  
# Checks the position and should end the game when the players score reaches 70, 
# however this code immediately assigns pos as 70 if pos is currently below 70, 
# ending the game. 
  
  if (pos < 70) {
    pos <- 70
    return(pos)
  }
}


#The fixed and adapted version of this code is as follows:

play_mc <- function(start_pos = 0, slippery_squares = c(9, 18, 27, 36, 45, 54, 63), climbing_squares = c(16, 32, 48, 64)) {
  die <- sample.int(6, 1)
  pos <- start_pos

  pos <- pos + die

# What has been changed here is an if statement to compute that if the score is  
# greater than or equal to 70 game ends. 
  
  if (pos >= 70) {  
    pos <- 70
    return(pos)
    }
  
#Adaption of function to include slippery squares & climbing squares.  

# Checks if the player's current position is one of the slippery squares, defined
# in function signature.
  if (pos %in% slippery_squares) {
    
# Randomly chooses how many squares the player will slide backward out of 3.
    slide <- sample(1:3, 1)
    
# Moves the player backwards by that amount, without allowing players position to go below 0.
    pos <- max(0, pos - slide)
  }

# Checks if the player's current position is one of the climbing squares also 
# defined in function signature.
  else if (pos %in% climbing_squares) {
    
# if so, moves the player forward by 5 squares, without exceeding 70.
    pos <- min(70, pos + 5)
  }
  
# Returns position
  return(pos)
}

play_mc()
```

```{r}

# Q2)

#Calls play_mc_full function

play_mc_full()

# Q3) 

#Assigns all output from simulate_climbs function to variable called results_sc.
results_sc <- simulate_climbs()

#Plots a histogram of turns vector that stores values of the amount of turns taken to finish the game
my_hist <- hist(results_sc$turns_vector,
# Labeling of axis and title for plot.
     main = "Histogram showing the frequency of turns taken to finish each game",
     xlab = "Number of turns",
     ylab = "Frequency")

# Q4) 

#Assigns all output from simulate_pos_at function to variable called results_sp.
results_sp <- simulate_pos_at()

#Plots a histogram of the position_at_n_turns that stores the values 
# of the positions of players after 10 turns taken.
n_turn_hist <- hist(results_sp$positions_at_n_turns,
# Labeling of axis and title for plot.
     main = "Histogram showing positions at 10 turns taken",
     xlab = "Position at 10 turns",
     ylab = "Frequency")


#Assigns all output from simulate_pos_at function to variable called results_sp_new_values.
results_sp_new_values <- simulate_pos_at(slip_back_range = 4:8, climb_up_amount = 2)

#Plots a histogram of the position_at_n_turns with new slip back range and climb
# up amount and shows the positions of players after 10 turns taken.
new_values_hist <- hist(results_sp_new_values$positions_at_n_turns,
# Labeling of axis and title for plot.
     main = "Histogram showing positions at 10 turns with updated values",
     xlab = "Position at 10 turns",
     ylab = "Frequency")

print(results_sp$turns_stats)
print(results_sp_new_values$turns_stats)

# For the first simulation we have a mean position of around 32 at 10 turns, whereas after changing the values and re-running the simulations with the updated slip back range and climb up amount, we have a mean position of around 27. This is expected based on the changes as the position is slipping back further and climbing up less when landing on slippery and climbing squares. This shift is demonstrated in the histograms shown.
```



```{r}

# Question 5

#Incorrect code line: #load("data/assess_data_1225.Rdata")
#Load() function takes an .Rdata file and loads it to current environment.
#Doesnt work as data file is not located in directory called data.
load("assess_data_1225.Rdata") #Correct line

# Density() function calculates a density curve for values of male BMI column 
#and assigns it to the variable dens_m.
dens_m <- density(bmi_data$bmi[bmi_data$sex == "male"])

#Calculates a density curve for values of female BMI column and assigns it to the variable dens_f.
dens_f <- density(bmi_data$bmi[bmi_data$sex == "female"]) 

#plot(dens_f, col = "green")
#lines(dens_m) Added labels & title code for clarity. 
#plot() function takes female density curve data and creates a graph, coloured green.
plot(dens_f, col = "green", main = "BMI Density Plot of Male and Female Participants", xlab = "BMI", ylab = "Density") 

lines(dens_m)#Adds male density curve data to the existing one

#Added code to make a legend to identify each line on density curve plot.
legend("topright", c("Female", "Male"), col = c("green", "black"), lty=1) 

#Assigns an index vector of value 1:200 to variable idx. 
idx  <- 1:200 

# prcomp() runs principle component analysis on a subset of data defined by idx,
#removes columns 1 & 9 from the subset, then assigns to variable pca_res.
pca_res <- prcomp(bmi_data[idx, -c(1, 9)], center = TRUE, scale = TRUE) 

#Colour by sex
colours <- ifelse(bmi_data$sex== "male", "blue", "pink") 

#Plots the first two principle components of pca_res
plot(pca_res$x[, 1], pca_res$x[, 2], col = colours, pch = 19, main = "Principal Component Analysis of BMI data subset", xlab = "Principle component 1", ylab = "Principle component 2")   

# out <- lm(bmi[idx] ~ age[idx]) Line was broken - did not reference data frame, fixed code includes this.
out <- lm(bmi_data$bmi[idx] ~ bmi_data$age[idx])  #lm() fits linear regression model, bmi as the response, age as the predictor on data subset defined by idx.

#Summary function provides a statistical overview of the linear regression model stored in variable out.
summary(out) 

# This line extracts p-value for the age coefficient from linear regression model and assigns it to variable p_val.
p_val <- summary(out)$coefficients[2, 4] 
p_val


```

Question 6)

```{r}
#This code produces a summary of the mean BMI values separated by sex. 
#Both averages are far beyond the normal range, indicating implausible values 
#and quality issues within the data set.

aggregate(bmi ~ sex, data = bmi_data, mean)
summary(bmi_data$bmi)


```

```{r}

# Histogram plot to visualise the incorrect BMI column before corrections


ggplot(bmi_data, aes(x = bmi)) +
  geom_histogram(fill = "darkblue", color = "black") +
#Code includes addition of dashed threshold lines to highlight the normal range of BMI values.
  geom_vline(xintercept = c(20, 30), linetype = "dashed", color = "black") +
  labs(title = "BMI Distribution of Participants", x = "BMI", y = "Count")

# Distribution is severely skewed, suggesting extreme values are pulling it to
# one side and not within normal ranges dashed lines added to visualise where normal
# BMI values should appear.

```

```{r}
# Code to calculate the true BMI values of each participant, new column added to dataframe bmi_data containing the true bmi data.
library(dplyr)
bmi_data <- bmi_data %>%
#BMI calculated using weight in kilograms divided by the square of height in meters.
  mutate(Correct_bmi = weight / ((height / 100)^2))

#Plots histogram of all the corrected BMI values and shows the histogram peak falls within the
# dashed lines showing expected range.
ggplot(bmi_data, aes(x = Correct_bmi)) +
  geom_histogram(fill = "darkblue", color = "black") +
  geom_vline(xintercept = c(20, 30), linetype = "dashed", color = "black") +
# Labelling of axis and title for plot.
  labs(title = "Distribution of Corrected BMI", x = "True BMI of Male and Female Participants", y = "Number of Participants")

# Plots a boxplot to show the corrected bmi for both sexes. Males here show a higher 
# median BMI although this is physiologically normal and still remains in normal range

ggplot(bmi_data, aes(sex, Correct_bmi, fill = sex)) +
  geom_boxplot() +
# Boxes coloured by sex to identify separate sexes.
  scale_fill_manual(values = c(female = "pink", male = "blue")) + 
#Code includes addition of dashed threshold lines to highlight the normal range of BMI values.
  geom_hline(yintercept = c(20, 30), linetype = "dashed") +
# Labelling of axis and title for plot.
  labs(title = "Boxplot of Corrected BMI", y = "True BMI", x = "Sex") 


```

```{r}

# Mean corrected BMI is now in the normal range, but min/max values remain abnormal and will require further QC, including remvoal of NAs.

# Boxplot shows corrected BMI medians in the expected range, but remaining outliers suggest more QC is needed.

# Summary of corrected BMI data, median/mean show values fall in normal range now.
aggregate(Correct_bmi ~ sex, data = bmi_data, mean)
summary(bmi_data$Correct_bmi)


```


```{r}

#PCA plot to visualise outliers using clean numerical data

pca_input <- bmi_data[, -c(1, 9)]
pca_input <- pca_input[sapply(pca_input, is.numeric)]

#Rows with NA or infinite values removed 
pca_input <- data.frame(lapply(pca_input, function(x) replace(x, !is.finite(x), NA)))
pca_input <- na.omit(pca_input)

p <- prcomp(scale(pca_input))
pc_dataframe <- as.data.frame(p$x)
var_explained <- round(100 * p$sdev^2 / sum(p$sdev^2), 1)

# Outliers computed using Tukey’s rule for outliers the aim is to set up a fence outside of Q1 and Q3, these are highlighted later in the code as outliers.

# IQR calculated for PC1
Q1_PC1 <- quantile(pc_dataframe$PC1, 0.25)
Q3_PC1 <- quantile(pc_dataframe$PC1, 0.75)
IQR_PC1 <- Q3_PC1 - Q1_PC1
lower_PC1 <- Q1_PC1 - 1.5 * IQR_PC1
upper_PC1 <- Q3_PC1 + 1.5 * IQR_PC1


# IQR calculated for PC2
Q1_PC2 <- quantile(pc_dataframe$PC2, 0.25)
Q3_PC2 <- quantile(pc_dataframe$PC2, 0.75)
IQR_PC2 <- Q3_PC2 - Q1_PC2
lower_PC2 <- Q1_PC2 - 1.5 * IQR_PC2
upper_PC2 <- Q3_PC2 + 1.5 * IQR_PC2


# Outliers identified using IQR values
outliers <- (pc_dataframe$PC1 < lower_PC1 | pc_dataframe$PC1 > upper_PC1) |
            (pc_dataframe$PC2 < lower_PC2 | pc_dataframe$PC2 > upper_PC2)


colours <- ifelse(bmi_data$sex == "male", "blue", "pink")  # Colour coding by sex

#Plot PCA
plot(pc_dataframe$PC1, pc_dataframe$PC2,
     pch = 19, cex = 0.3, 
     col = colours,      
     xlab = paste0("PC1 (", var_explained[1], "%)"),
     ylab = paste0("PC2 (", var_explained[2], "%)"),
     main = "PCA of BMI Data with Outliers Highlighted")

#Plots points identified as outliers 
points(pc_dataframe$PC1[outliers], pc_dataframe$PC2[outliers],
       col = "red",      # Outliers highlighted in red
       pch = 19, 
       cex = 0.5)

# Legend to colour code male points, female points and outliers. 
legend("topright", legend = c("Male", "Female", "Outlier"),
      col = c("blue", "pink", "red"), pch = 19) 

#PCA plot shows two clear clusters containing data points from males and females,
# red outliers show extreme values remain in individuals data within the dataset.
# PC1 explains 55.6% of the variance across all numeric variables a further 26.8%
# is explained by PC2. 84.4% is therefore explained by PC1 and PC2, allowing trends
# and clusters to accurately be shown in the PCA plot.

```


```{r}
# Summary of cleaned bmi data, shows all NA values removed, outliers have been removed, min/max values are now not implausible. This function also replaces original bmi column.
cleaned_bmi_data <- update_bmi(bmi_data)
summary(cleaned_bmi_data)
```
Q7 Using simple linear regression to check how well bmi (covariate) can predict body_fat (response). Check assumptions of the simple linear regression model and also plot the residuals. Comment on the results of your analysis.

Comment on the residuals, what is the shape and what does this tell you about the model fit and any other covariates that should be included?

```{r}
# Question 7)

library(ggplot2)
library(reshape2)
library(bbmle)

#Fits a simple linear regression model
lm_model <- lm(body_fat ~ Correct_bmi, data = cleaned_bmi_data)
summary(lm_model) #Gives a summary of the model

#Plots linear regression model producing four diagnostic plots.
plot(lm_model)

# Residuals vs Fitted plot shows the residuals are scattered around 0, meaning linearity assumption is reasonably met, although funneling at the end indicates heteroscedasticity so residuals have non constant variance, the plot also highlights outliers. Variance is also shown to increases as fitted values increase. R squared value of 0.2956 also shows that BMI explains only 29.6% of the variation in body fat meaning this is not a great model.


#Q-Q plot is checking for distributional assumptions and shows both positive and negative tails deviate from the theoretical distribution, the heavy tail indicates that the residuals have more extreme values than expected under normality, while the lighter tail means there are less extreme residuals than would be expected under normality. This is most likely influenced by the outliers and the lack of other covariates influence being accounted for.

#Linear regression model plot:

#To plot linear regression model for body fat vs BMI.
lr_plot <- ggplot(cleaned_bmi_data, aes(x = Correct_bmi, y = body_fat)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "blue") +  # Adds a regression line 
  labs(
    title = "Simple Linear Regression model: Body Fat vs BMI",
    x = "Predictor: BMI",
    y = "Response: Body Fat")

print(lr_plot)

# This plot shows the regression line on the scatter plot of BMI vs Bodyfat, there 
# is shown to be a linear correlation between the predictor and the response although
# BMI alone is not shown to be a reliable predictor of bodyfat.

residuals_df1 <-
data.frame(resid = residuals(lm_model))


# To plot histogram of residuals for body fat vs BMI linear model.
ggplot(residuals_df1, aes(x = resid)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 30, 
                 fill = "darkblue", 
                 color = "black") +
  
# Code to overlay a normal curve on the histogram in light blue
  stat_function(fun = dnorm,
                args = list(mean = mean(residuals_df1$resid, na.rm = TRUE),
                            sd = sd(residuals_df1$resid, na.rm = TRUE)),
                color = "lightblue", 
                size = 1) +
#Labels added for title and x,y values
  labs(title = "Histogram of Residuals with Fitted Normal Curve",
       x = "Residuals",
       y = "Density")


# Histogram lacks bell-shaped curve for normal distribution and instead indicates a bimodal distribution shown but the two separate peaks and trough around 0. This likely means there is a presence of two separate groups where there is a concentration of data points clustering there. This means it is likely that sex is one of the main covariates that should be incorporated, based on the knowledge that males tend to have a higher BMI than females shown in previous statistics, therefore this could be included into this model for an improved model fit.
```



```{r}

# This code fits a multiple linear regression model predicting body fat using age, BAI, body density, weight, height, and corrected BMI as predictors with body fat as the response. 
lm_bodyfat_covariants <- lm(body_fat ~ age + bai + density + weight + height + Correct_bmi, 
                 data = cleaned_bmi_data)
summary(lm_bodyfat_covariants)
# Summary() function displays coefficient estimates, standard errors, R-squared, 
# and statistical significance for each predictor. Summary of all predictors shows
# R-squared value of 0.9961, means that 99.6% of the variation in body fat is 
# explained by the 6 covariates. Indicates the model fit is very strong and a good predictive model.
# Based on the absolute t-value which shows how much each predictor contributes to the 
# prediction of body fat, accounting for the variance of the predictor, we can see that 
# density is by far the dominant predictor of body fat. 


# plot() function produces four plots of residuals for analysis
plot(lm_bodyfat_covariants)

# Residuals vs Fitted plot shows points are symmetrically distributed, clustering 
# towards the middle of the plot. This suggests that the residuals have a generally constant 
# variance and that the model does not show obvious patterns indicating non-linearity.
# There are two outliers in the Residuals vs Fitted plot, that appear far from the 
# bulk of data, these will be having an influence on the model's estimates. However, 
# since quality control has already been used to clean the data it is likely these 
# just represent natural variability in the population.

# Q-Q plot further shows the normal distribution of residuals, the residuals are 
#closely plotted against the expected quantiles of a normal distribution. 
# Slight deviations from the norm however occur on the positive and negative tail,
# however generally indicates a normal distribution.

# Creates a data frame containing residuals extracted from regression model.
residuals_df2 <-
data.frame(resid = residuals(lm_bodyfat_covariants))


# Creates a histogram plot of all the predictors to examine the performance of all covariates in predicting body fat.
ggplot(residuals_df2, aes(x = resid)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 30, 
                 fill = "darkblue", 
                 color = "black") +
  
# Code to overlay a normal curve on the histogram in light blue
  stat_function(fun = dnorm,
                args = list(mean = mean(residuals_df2$resid, na.rm = TRUE),
                            sd = sd(residuals_df2$resid, na.rm = TRUE)),
                color = "lightblue", 
                size = 1) +
#Labels added for title and x,y values
  labs(title = "Histogram of Residuals with Fitted Normal Curve",
       x = "Residuals",
       y = "Density")


# Residuals histogram shows a normal distribution and closely follows the normal curve, with clustering around the mean and median. Supports R-squared value of 0.9961, meaning this is a strong model. 
```

